# Comparisons
### Load results
#### Microarray: GSE5145
```{r}
gse5145 <- read.csv("results/GSE5145/array5145_genes.csv")
gse5145_gse <- read.csv("results/GSE5145/array5145_genesets.csv")
```

#### Microarray: GSE145483

```{r}
gse145483 <- read.csv("results/GSE145483/array145483_genes.csv")
gse145483_gsa <- read.csv("results/GSE145483/array145483_genesets.csv")
```

#### RNAseq: GSE69284

```{r}
gse69284_2.5h <- read.csv("results/GSE69284/genes2.5h.csv")
gse69284_4h <- read.csv("results/GSE69284/genes4h.csv")
gse69284_24h <- read.csv("results/GSE69284/genes24h.csv")
gse69284_2.5h_gsa <- read.csv("results/GSE69284/gsa_1.csv")
gse69284_4h_gsa <- read.csv("results/GSE69284/gsa_2.csv")
gse69284_24h_gsa <- read.csv("results/GSE69284/gsa_3.csv")
```

#### RNAseq: GSE189984

```{r}
gse189984_4 <- read.csv("results/GSE189984/genes4h.csv")
gse189984_8 <- read.csv("results/GSE189984/genes8h.csv")
gse189984_24 <- read.csv("results/GSE189984/genes24h.csv")
gse189984_48 <- read.csv("results/GSE189984/genes48h.csv")
```

### Common genes
#### Venn diagram
```{r}
#install.packages("ggvenn", update=FALSE)              
library(ggvenn)

# filter na values
gse189984_48 <- gse189984_48[!is.na(gse189984_48$gene_symbol),]
gse69284_24h <- gse69284_24h[!is.na(gse69284_24h$gene_symbol),]
gse5145 <- gse5145[!is.na(gse5145$Gene.symbol),]
gse145483 <- gse145483[!is.na(gse145483$Gene),]

# List of items
x <- list(gse5145_TMA = gse5145$Gene.symbol, gse69284_RNAseq = gse69284_24h[gse69284_24h$FDR < 0.05,]$gene_symbol, gse145483_MMA = gse145483[gse145483$adj.P.Val < 0.05,]$Gene, gse189984_RNAseq = gse189984_48[gse189984_48$FDR < 0.05,]$gene_symbol)

ggvenn(x)
```

Quite a lot of common genes between GSE69284 and GSE145483: explainable by the fact that those were both performed on monocytic cell lines.
This shows the genes that are regulated by VitD can be very different for different cell types.

#### Get the common genes:
##### All the data sets
```{r}
Reduce(intersect, list(gse5145$Gene.symbol,
          gse69284_24h[gse69284_24h$FDR < 0.05,]$gene_symbol,
          gse145483[gse145483$adj.P.Val < 0.05,]$Gene,
          gse189984_48[gse189984_48$FDR < 0.05,]$gene_symbol))
```

##### Between monocytic data sets

Same cell types -> we expect functions associated with Vit D for this specific cell type

```{r}
common <- Reduce(intersect, list(gse145483[gse145483$adj.P.Val < 0.05,]$Gene,
                       gse69284_24h[gse69284_24h$FDR < 0.05,]$gene_symbol))
```

Perform GSA on this common set of genes:

```{r}
library(limma)

GRCh38 <-  useEnsembl(biomart="genes", host="https://www.ensembl.org",
                      dataset="hsapiens_gene_ensembl")

# attributes
atr <- listAttributes(GRCh38)
idmap <- getBM(attributes = c('entrezgene_id',
                             'external_gene_name'), mart = GRCh38)

# map gene symbols to entez ids
idmap <- idmap[!duplicated(idmap$external_gene_name),]
idmap <- idmap[idmap$external_gene_name %in% common,]
entrez <- idmap$entrezgene_id

res <- kegga(de=entrez,
             species.KEGG="hsa")
res <- res[order(res$N),]
    
res$FDR.DE <- p.adjust(res$P.DE, n=nrow(res), method="BH")
```

Analyse results:

```{r}
head(res[order(res$P.DE),])
```

TODO: check the meaning behind these?

- role of VitaminD in Atherosclerosis (TODO: ref)?
