---
title: "Group 11 - VitD-OmicsAnalysis"
author: "Robbe & Simon"
date: "`r Sys.Date()`"
output: html_document
---

# Introduction
## Project Overview
Group number: 11
Group memebers: Robbe Claeys & Simon Bil  

**Research question:** An omics-wide analysis of Vitamin D difieciency.

## Data Overview
**Paper links:**

RNA microarray

- https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4030167/
  - https://doi.org/10.1152/physiolgenomics.00134.2006
  - https://doi.org/10.1096/fj.10-172577

ChIP & RNA seq

- https://www.nature.com/articles/s41598-021-86032-5

**Download links:**

RNA microarray (GSE5145)

- https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE5145

RNA microarray (GSE22523)

- https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE22523

RNA seq (GSE69284)

- https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE69284

ChIP seq (GSE27438)

- https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE27438

```{r}
table_general_info <- rbind(c("GSE5145", "GSE22523", "GSE69284", "GSE27438"), c(6, 8, 18, 5), c(6, 8, 18, 5))

colnames(table_general_info) <- c("Microarray Data", "Microarray Data", "RNAseq Data", "ChIPseq Data")

rownames(table_general_info) <- c("Accesion ID", "Number of Samples (original)", "Number of Samples (analysed)")

table_general_info
```

---

Loading all libraries

```{r echo = T, results = 'hide', error=FALSE, warning=FALSE, message=FALSE}
library(htmltools)

# for microarrays
library(oligo)
library(GEOquery)
library(arrayQualityMetrics)
library(limma)

# for RNA seq
library(tximport)
library(dplyr)

# for ChIP seq

```

# Microarray Data - GSE5145

## General info

This dataset was generated in 2007 and aimed to identify the genes regulated by Vit D in primary bronchial smooth muscle cells. Cells were treated for 24 hours with 100 nM of Vit D3. The experiment was done in triplicate form (a total of 6 samples were analyzed). All samples of this study were used for our data analysis.

**Microarray type:** Affymetrix Human Genome U133 Plus 2.0 Array [HG-U133_Plus_2]

## Intensity values

Load the data.

```{r echo = T, results = 'hide', error=FALSE, warning=FALSE, message=FALSE}
exonCELs <- list.celfiles("./data/microarray/GSE5145/")
arrayRaw <- read.celfiles(paste(rep("./data/microarray/GSE5145/", length(exonCELs)), exonCELs, sep = ""))
```

Get dimensions and general overview of the data. The first three `.CEL` files are the Vit D deficient samples, the last three are the control samples.

```{r}
dim(exprs(arrayRaw))
head(exprs(arrayRaw))
```

Extraction of the expression values using the robust multichip average method (RMA) and summarization of the array data.

```{r}
arrayRma = rma(arrayRaw)
dim(exprs(arrayRma))
head(exprs(arrayRma))
```

## Annotation

We will be investigating differential expression between Vit D deficient and normal primary bronchial smooth muscle cell samples. No other confounding factors are present in the sample annotation. All primary bronchial smooth muscle cells were derived from the same tissue sample of a 31 year old Hispanic male.

```{r error=FALSE, warning=FALSE, message=FALSE}
annot <- getGEO(filename = "./data/annotation/GSE5145/GSE5145_family.soft")

IDs <- annot@header$sample_id
titles <- c()

for (id in IDs) {
  title <- annot@gsms[[id]]@header[["title"]]
  titles <- append(titles, title)
}
data.frame(IDs, titles)
```

## Explorative quality evaluation

Quality of the array determined with the arrayQualityMetrics package.

```{r}
arrayQualityMetrics(expressionset = arrayRma,
                    outdir = "./stats/Microarray/GSE5145/",
                    force = TRUE)
```

From this analysis we can conclude that the overall quality of the arrays is quite good. However, sample 6 shows signs of being an outlier and the graph on figure 7 (Standard deviation versus rank of the mean) shows an upward trend, which is symptomatic of a saturation of the intensities.

## Differential expression analysis

We will perform LIMMA to look for differential expressed genes.

```{r}
# design matrix
design <- model.matrix(~ 0 + factor(c(1,1,1,2,2,2)))
colnames(design) <- c("VitD", "Control")
fit <- lmFit(arrayRma, design)
```

To make a pair-wise comparison between the two groups the appropriate contrast matrix is created.

```{r}
contrast.matrix <- makeContrasts(Control-VitD, levels = design)
fit2 <- contrasts.fit(fit, contrast.matrix)
fit2 <- eBayes(fit2)
```

A list of top genes differential expressed in group2 (Control) versus group1 (VitD) can now be obtained.

The best probe (203887_s_at) is expressed 10^3 times less in the Control group (log fold change). All 500 probes in this list have significant adjusted p-values.

```{r}
topNumber <- 500
table <- topTable(fit2, coef=1, number=topNumber, adjust="BH") # FDR correction of p-values
table
```

## Summarization at the gene level

Note that we are now working at the level of probe sets and not at level of genes per se. Whereas many genes are only represented by a single probe set on a micro array, for some genes multiple probe sets are present

```{r}
# input data (all of equal length)
allProbeIDs <- annot@gpls[["GPL570"]]@dataTable@table[["ID"]]
allGeneIDs <- annot@gpls[["GPL570"]]@dataTable@table[["ENTREZ_GENE_ID"]]
allGeneSym <- annot@gpls[["GPL570"]]@dataTable@table[["Gene Symbol"]]
allGeneOnt <- annot@gpls[["GPL570"]]@dataTable@table[["Gene Ontology Biological Process"]]
```

```{r warning=FALSE}
# desired data
geneIDs <- c()
geneSym <- c()
logFCs <- c()
geneOnt <- c()

for (topIndex in 1:topNumber) {
  probeID <- rownames(table)[topIndex]
  annotIndex <- grep(probeID, allProbeIDs)
  
  if (!(allGeneIDs[annotIndex] %in% geneIDs) & (probeID %in% allProbeIDs)) {
    geneIDs <- append(geneIDs, allGeneIDs[annotIndex[1]])
    geneSym <- append(geneSym, allGeneSym[annotIndex[1]])
    logFCs <- append(logFCs, table[[1]][topIndex])
    geneOnt <- append(geneOnt, allGeneOnt[annotIndex[1]])
  } 
}

df <- data.frame(Entrez.gene.ID = geneIDs, Gene.symbol = geneSym, Fold.Change = logFCs, Gene.ontology = geneOnt)
df
```

The top 500 most significant probes could be mapped to 330 individual genes, which were mostly upregulated after Vit D treatment. Many of these genes have a role in high-level signal transduction (cell cyle, metabolism, differentiation). This is a first indication that Vit D is an important nutrient with pleiotropic regulatory effects in the cell.

When comparing these findings to those reported in the original paper, the same upregulated genes were found.

---

# Microarray Data - GSE22523
## General info
This dataset (E-GEOD-22523) was generated in 2010 and aimed to identify genes regulated by Vit D in B cells. A gene expression microarray analysis in HVDRR EBV-transformed B-cells and control cells was performed in the presence or absence of vitamin D. This study was performed on 8 samples (4 cases & 4 controls), all were used for our data analysis.

**Microarray type:** Affymetrix Human Gene 1.0 ST Array [HuGene-1_0-st]

## Intensity values
Load the data
```{r echo = T, results = 'hide', error=FALSE, warning=FALSE, message=FALSE}
exonCELs <- list.celfiles("./data/microarray/GSE22523/")
arrayRaw <- read.celfiles(paste(rep("./data/microarray/GSE22523/", length(exonCELs)), exonCELs, sep = ""))
```

The first 4 `.CEL` files contain the data for the hereditary vitamin D-resistant rickets (HVDRR) cells, of which the first two are biological replicates and the last two were treated with Vit D. The last 4 `.CEL` files contain the data for the Wild-type (WT) cell, again, of which the first two are biological replicates and the last two were treated with Vit D.

Get dimensions and general overview of the data. 
```{r}
dim(exprs(arrayRaw))
head(exprs(arrayRaw))
```

Summarization of the array data.
```{r}
arrayRma = rma(arrayRaw)
dim(exprs(arrayRma))
head(exprs(arrayRma))
```

## Annotation
We will be investigating differential expression between VitD treated & untreated cel cultures. Confounding factors: there are 4 wild-type and 4 HVDRR cells.

```{r error=FALSE, warning=FALSE, message=FALSE}
annot <- getGEO(filename = "./data/annotation/GSE22523/GSE22523_family.soft")

IDs <- annot@header$sample_id
titles <- c()

for (id in IDs) {
  title <- annot@gsms[[id]]@header[["title"]]
  titles <- append(titles, title)
}
data.frame(IDs, titles)
```

## Explorative quality evaluation

Quality of the array determined with the arrayQualityMetrics package.

```{r warning=FALSE}
arrayQualityMetrics(expressionset = arrayRma,
                    outdir = "./stats/Microarray/GSE22523/",
                    force = TRUE)
```

From this analysis we can conclude that the overall quality of the arrays is very good. No problematic outliers or variance trends were observed.

## Differential expression analysis

We will perform LIMMA to look for differential expressed genes.

```{r}
design <- model.matrix(~ factor(c(1,1,2,2,1,1,2,2)) + 
    factor(c(1,1,1,1,2,2,2,2)))
colnames(design) <- c("Intercept", "Treated", "Celltype")
fit <- lmFit(arrayRma, design)
fit <- eBayes(fit)
```

A list of top differential expressed genes in the treated groups versus the control groups can now be obtained by evaluating the first coefficient. The cell type coefficient did not yield significant findings, indicating that cell type did not play a deciding role in the differential expression.

The best probe (7894912) is expressed 10^-0.9 times less in the Control group (log fold change). Notably, no significant findings were retrieved after FDR correction of the p-values.

```{r}
topNumber <- 500
table <- topTable(fit, coef = 2, number = topNumber, adjust = "BH") # FDR correction of p-values
table
```

## Summarization at the gene level

For annotation, the different probe sets are mapped to individual genes present in the .soft file.

```{r}
# input data (all of equal length)
allProbeIDs <- annot@gpls[["GPL6244"]]@dataTable@table[["ID"]]
allGeneIDs <- annot@gpls[["GPL6244"]]@dataTable@table[["gene_assignment"]]
```

```{r}
# desired data
geneIDs <- c()
logFCs <- c()

for (topIndex in 1:topNumber) {
  probeID <- rownames(table)[topIndex]
  annotIndex <- grep(probeID, allProbeIDs)
  
  if (!(allGeneIDs[annotIndex] %in% geneIDs) & (probeID %in% allProbeIDs)) {
    geneIDs <- append(geneIDs, allGeneIDs[annotIndex[1]])
    logFCs <- append(logFCs, table[[1]][topIndex])
  } 
}

df <- data.frame(Entrez.gene.ID = geneIDs, Fold.Change = logFCs)
df
```

The top 500 most significant probes could be mapped to 235 individual genes. Again, we see many general signal transduction factors (e.g. elongation factor) as well as general-purpose factors such as the mitochondrial rRNA transcripts.

When comparing these findings to those reported in the original paper, only a few matches were found. This is because the paper was specifically investigating transcripts with a role in immune signaling, thus only these were evaluated.

---

TODO: include other file?

---

# ChIPseq Data - GSE27438

## General info
This dataset (GSE27438) was generated in 2011 and aimed to identify genomic locations bound by the vitamin D receptor (VDR).
ChIP-seq for the vitamin D receptor (VDR) was performed in the monocytic THP-1 cells after stimulation with the natural VDR ligand 1,25-dihydroxyvitamin D3 (1,25(OH)2D3) or with vehicle for 4h before ChIP assay.

- Read length: 36bp
- Single/paired end sequencing: single end sequencing
- Platform used: Illumina Genome Analyzer II

## Quality control on fastQ files

```{r, echo=FALSE, results='asis'}
htmltools::includeHTML("./stats/ChIPseq/multiqc_report.html")
```

- Watch out bcause there are two samples with lots of reads (fastQC sequence counts)

## Mapping statistics

> We tried to perform mapping ourselves, but failed at a certain point.
> For this reason we did not have time to finish peak data and annotation on the ChIPseq data.

## Annotation

```{r}
gse27438_annot <- read.csv("./data/annotation/GSE27438/SraRunTable.txt")
print(gse27438_annot[c("Run", "Treatment")])
```

